---
title: "Take Home Exam 1"
output: 
  flexdashboard::flex_dashboard:
    orientation: columns
    vertical_layout: fill
    theme: cosmo
---

```{r setup, include=FALSE}
library(flexdashboard)
library(tidyverse)
```
Scatterplot
===

**Null Hypothesis (H₀):** Crocodile weight does not depend on length  
**Alternative Hypothesis (H₁):** Crocodile weight depends on length  
**Test Statistic:** 49.60  
**P-Value:** 2e-16, reject null hypothesis as 2e-16<0.05

Using the regression model to explain the weight of the a crocodile based on length is a much better way to explain the data than just using the mean. This is because using the regression model accounts for a trend in the data while the mean completely ignores any trends.


Column {data-width = 100}
---
```{r include=FALSE}
crocodile <- read.csv("data/crocodile_dataset.csv")
glimpse (crocodile)

df <- crocodile %>% 
  filter(Genus== "Mecistops",
         Sex=="Male")

model <- lm(crocodile$Observed.Weight..kg.~crocodile$Observed.Length..m.)
summary(model)

confint(model, level = 0.95)
```

Column {data-width = 450}
---
```{r}
plot(crocodile$Observed.Weight..kg.~crocodile$Observed.Length..m.,
     data = df,
     xlab = "Observed Length (meters)",
     ylab = "Observed Weight (kilograms)",
     main = "Relationship Between Crocodile Length and Weight")
```

Confidence and Prediction Intervals
===

**Confidence Interval:** We are 95% confident that for every 1 meter increase in length, the average crocodile weight increase by between 129.3 kilograms and 139.95 kilograms

Column {data-width=400}
---
```{r include = FALSE}
confint(model, level = 0.95)

newdata <- data.frame(Observed.Length..m. = c(3.5, 4,5))

predictions <- predict(model, newdata, interval = "prediction", level = 0.95)
print(predictions)

```

```{r}
plot(predictions,
     xlab = "lower",
     ylab = "fit",
     main = "95% Prediction Interval")
```

Evaluation
===

The model provides from predictive ability but the confidence intervals suggest that the model doesn't fully show the relationship between length and weight due to limitations in this linear relationship. For example, outliers, a smaller sample size, and a slightly nonlinear overall trend.

Concerns when applying a linear regression model:<br>
  -  Typical weight to length in animals is usually not exactly linear but following a more nonlinear trend <br>
  - Any outliers or extremities could be ignored through a linear regression model  <br>
  - If the data set is small, the model may not be representative of all data values